Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-29 code/ms2_bias_niche.Rmd
almost done ms2...

ggsave("../output/pcoa_aggregate.png")
```

@@

# Isolate genus-level, then pool within tongue.pair and site
niche_overlap_tongues <- niche_overlap %>%
  group_by(site, level, tongue.pair, elev.mean, management) %>%
  summarize(mean.niche.overlap = mean(niche.overlap)) %>%
  filter(level == "genus")

@@

# Plot intercept effect of tongue.diff
plot(pterm(niche_gam_20, 1))
```

>>>>

ggsave("../output/pcoa_aggregate.png")
ggsave("../output/pcoa_aggregate.pdf")

@@

# Isolate genus-level, then pool within tongue.pair and site
niche_overlap_tongues <- niche_overlap %>%
  group_by(site, level, tongue.pair, elev.mean, management) %>%
  summarize(mean.niche.overlap = mean(niche.overlap)) %>%
  filter(level == "genus") %>%
  mutate(tongue.pair = factor(tongue.pair, levels = c("short:long", "short:med", "short:short",
                                                      "med:long", "med:med", "long:long")))

@@

# Plot intercept effect of tongue.diff
plot(pterm(niche_gam_21, 1))
```

### 2021-03-29 ms_2/ms_2_v2.Rmd
almost done ms2...

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the degree of morphological fittedness between bumble bees and their floral hosts, which consists principally in the matching of bumble bee tongue length to floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], which leads to an association between tongue length and floral choice in bumble bee communities [e.g. @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. This association between tongue length and floral niche is imperfect, though, and there can be considerable overlap in floral resource use between bumble bees of different tongue lengths [e.g. @Ranta1980-ok; @Goulson2008-oe].

@@

To focus on genus-level niche overlap as a function of tongue length class, we generated a binary tongue-difference variable for each species pair: "same" for species belonging to the same tongue length class and "different" for species belonging to different tongue length classes. We then took the mean genus-level niche overlap across all species pairs within tongue-difference class and site. Then, we constructed an HGAM of genus-level niche overlap using tongue-difference as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf.

@@

Morphotype-level diet overlap was nearly double genus-level diet overlap throughout the whole elevation gradient, with family-level diet overlap taking intermediate values. The fixed intercepts of morphotype- and genus-level diet overlap each differed significantly (in opposite directions) from the reference level of family-level diet overlap (p << 0.001). The conditional smooths of each level were significantly different from flat lines (p < 0.001) and generally mirrored one another in shape **(Figure 8)**. Diet overlap was highest at the low end of the elevation gradient, decreased steeply from 600-1200 m, and leveled off or declined weakly for the remainder of the gradient, though morphotype-level diet overlap exhibited a second peak just above 1500 m that was not evident in genus- or family-level diet overlap. Model 5 explained 74.2% of total deviance (adjusted R^2^ = 0.70). Whether a species pair belonged to the same or different tongue length classes had no effect on genus-level niche overlap, either in terms of intercept (p = 0.12) or shape of response to elevation **(Figure 9)**. Model 6 explained 40.0% of total deviance (adjusted R^2^ = 0.35).

@@

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show 95\\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_20_plot1.png")
```

@@

Any discussion of competition and niche partitioning among *Bombus* species must begin with the observation that bumble bees are unique among insect pollinators in the degree to which their foraging combines contrasting extremes of bias and flexibility. Coexistence between bumble bee species with similar tongue lengths hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. In addition to addressing the question of bumble bee species coexistence, our results suggest how patterns of competition may respond to climate and land use change along mountain slopes, an issue of immediate concern for the conservation of bumble bees and their floral mutualists [@Sirois-Delisle2018-vi; @Soroye2020-zb].

@@

 While biases can be explained partly in terms of morphological matching between tongue length and floral morphotype

 The rule is flexibility, and floral morphotypes do not constitute hard boundaries to bumble bee flower choice. Nevertheless, we demonstrate that for certain floral morphotypes --- funnel flower, bell flowers, flag flowers, and head flower --- visitation rate is affected strongly by tongue length.

In our study, lip, flag, and head flowers were by far the most strongly preferred when visitation was pooled across all bumble bee species, while disc, funnel, bell, and stalk-disc flowers were comparatively underrepresented. These findings are broadly consistent with the aggregate visitation patterns previously reported by @Schneller2014-ry in the central Alps at Vorarlberg and Neumayer1999-il ~50 km south of our study system, suggesting that our findings are at least regionally generalizeable. When considering these patterns from the perspective of pollination services to mountain flora, however, it is important to note that while bumble bees are the dominant Hymenopteran pollinators in our study system, Syrphid flies are also abundance flower visitors whose visitation is mainly restricted to the open flower morphotypes (e.g. disc flowers) that receive less attention from bumble bees [@Neumayer1999-il]. In the context of our research question, we have emphasized the differences in tongue length between bumble bee species, but all bumble bees have relatively long tongues compared to the overall pollinator community, and the process of niche partitioning *among* bumble bee species is nested within the broader process of niche partitioning *between* bumble bees and other flower visitors.

While the variation we observed in floral preference across tongue length classes largely comports with the established relationship between tongue length and corolla depth, variation in preference *within* tongue length classes agrees with studies demonstrating the insufficiency of simple morphological matching in explaining bumble bee foraging patterns [e.g. @Ranta1980-ok; @Goulson2008-oe]. Perhaps the most striking divergence of floral preference within tongue length class was the divisive effect of bell flowers on short-tongued bumble bees. *B. pyrenaeus*, *B. soroeensis*, *B. pratorum*, and *B. jonellus* visited bell flowers at high rates, while *B. terrestris/lucorum*, *B. psithyrus*, and *B. lapidarius* visited them relatively rarely. A similar split among the short-tongued species occurred with respect to head flowers, which were visited at high rates by *B. soroeensis* and *B. psithyrus* but at relatively low rates by other short-tongued species. There are no obvious morphological distinctions within short-tongued bumble bees that would account for these divergent affinities, and ecological drivers such as differences in spatial or temporal activity patterns are more likely explanations [@Sowig1989-ld; @Schneller2014-ry]. Nectar robbing is another mechanism by which the relationship between tongue length and corolla depth might be relaxed, but its influence in our data is unclear. Of the bumble bees in our study system, *B. terrestris/lucorum* and especially *B. wurflenii* are known to engage frequently in nectar robbing [@Brian1954-cy]. In the case of *B. wurflenii*, however, its morphotype affinities did not differ markedly from those of other medium-tongued species, and it may be that the condition of medium-tongue length alone confers enough to foraging flexibility in our study system to dampen the divergent effects of nectar robbing. Nectar robbing might, however, account for the deviation of *B. terrestris/lucorum* from most other short-tongued species with respect to flag flowers. With the exception of *B. lapidarius*, all short-tongued species besides *B. terrestris/lucorum* eschewed flag flowers, while *B. terrestris/lucorum* visited them at rates comparable to medium- and long-tongued species. Indeed, *B. terrestris/lucorum* was, on the whole, an eclectic forager, also demonstrating a relatively high affinity for the disc flowers that were largely ignored by other species. It should be noted, though, that an unusually high abundance of *B. terrestris/lucorum* queens was observed in 2012, but nest founding success was evidently low, leading to low *B. terrestris/lucorum* abundance later in the season, an anomaly that could account for the unusual foraging patterns of this species group in the pooled data. Together, these species-level idiosyncracies of preference within tongue length classes underscore the fact that factors other than the matching of tongue length and floral morphology --- such as affinities based on floral color, plant growth form, spatial distribution of floral patches, or habitat context [@Brian1957-dg; Harder1985-ls; @Sowig1989-ld] --- are at play in bumble bees' floral preferences, which may partly explain the coexistence of bumble bees of similar tongue lengths.

<!-- - Explain that traits such as color (really especially color) probably explain much of the variance in preference not explained by morphology [@Neumayer1999-il].  -->

@@

With respect to diet overlap among bumble bees, our finding that bumble bee diet ordination produces a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. Perhaps a more significant finding, though, is that community-mean diet overlap at the level of floral morphotype was consistently nearly twice as high as diet overlap at the level of floral genus. From a numerical perspective, this result is unsurprising simply because each morphotype comprises many genera. From an ecological perspective, though, this pattern suggests another answer to the question of how bumble bee species of similar tongue length can coexists. If tongue length primarily constrains bumble bee foraging at the coarse level of floral morphotype, floral diversity *within* morphotypes creates space for bumble bees with similar tongue lengths to avoid actual dietary overlap, thus mitigating interspecific competition. Importantly, this suggests that diversity within floral morphotypes may as important as morphotype diversity *per se* and overall floral abundance in facilitating the coexistence of bumble bee species, an insight which may have implications for the design of floral enhancements aimed at bumble bee conservation.

<!-- @Durieux2000-uk similarly ofund that while tongue length is correlated with floral preference, it is not correlated with diet similarity. But we did better stats! -->

@@

@@

<!-- Lip and head flowers have the highest nectar reward per inflorescence [@Neumayer1999-il]. -->
<!-- Interestingly, @Farnberger1996-xo in a localized study near Wien found rather different results. Lip flowers were still found to be the favorite overall, and head flowers were popular, too, but flag flowers were of secondary importance, and bell flowers were the most visited morphotype for humi, pasc, and prat. Pasc almost totally ignored bell flowers in our study. -->

>>>>

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the degree of morphological fittedness between bumble bees and their floral hosts, which consists principally in the matching of bumble bee tongue length to floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], which leads to an association between tongue length and floral choice in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. This association between tongue length and floral niche is imperfect, though, and there can be considerable overlap in floral resource use between bumble bees of different tongue lengths [e.g. @Ranta1980-ok; @Goulson2008-oe].

@@

<!-- To focus on genus-level niche overlap as a function of tongue length class, we generated a binary tongue-difference variable for each species pair: "same" for species belonging to the same tongue length class and "different" for species belonging to different tongue length classes. We then took the mean genus-level niche overlap across all species pairs within tongue-difference class and site. Then, we constructed an HGAM of genus-level niche overlap using tongue-difference as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf. -->

To focus on genus-level niche overlap as a function of tongue length class, we classified each species pair as a combination of "short:short", "short:medium", "short:long", "medium:medium", "medium:long", or "long:long". Within each site, we then took the mean genus-level niche overlap across all species pairs within each tongue-pair class. Then, we constructed an HGAM of genus-level niche overlap using tongue-pair as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf.

@@

Morphotype-level diet overlap was nearly double genus-level diet overlap throughout the whole elevation gradient, with family-level diet overlap taking intermediate values. The fixed intercepts of morphotype- and genus-level diet overlap each differed significantly (in opposite directions) from the reference level of family-level diet overlap (p << 0.001). The conditional smooths of each level were significantly different from flat lines (p < 0.001) and generally mirrored one another in shape **(Figure 8)**. Diet overlap was highest at the low end of the elevation gradient, decreased steeply from 600-1200 m, and leveled off or declined weakly for the remainder of the gradient, though morphotype-level diet overlap exhibited a second peak just above 1500 m that was not evident in genus- or family-level diet overlap. Model 5 explained 74.2% of total deviance (adjusted R^2^ = 0.70).

<!-- Whether a species pair belonged to the same or different tongue length classes had no effect on genus-level niche overlap, either in terms of intercept (p = 0.12) or shape of response to elevation **(Figure 9)**. Model 6 explained 40.0% of total deviance (adjusted R^2^ = 0.35). -->

We recommend caution, however, in interpreting the p-values of this model, since the pairwise comparison of group differences entails a degree of non-independence.

@@

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show 95\\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_21_plot1.png")
```

@@

@@

Any discussion of competition and niche partitioning among *Bombus* species must begin with the observation that bumble bees are unique among insect pollinators in the degree to which their foraging combines contrasting extremes of bias and flexibility. While the floral preferences of bumble bee species can be explained partly by morphological matching between tongue length and floral morphotype, floral morphotypes are not hard boundaries to flower choice [@Kugler1943-uy; @Brian1954-cy]. Both terms of this paradox are evident in our study.

Lip, flag, and head flowers were by far the most strongly preferred when visitation was pooled across all bumble bee species, while disc, funnel, bell, and stalk-disc flowers were comparatively underrepresented. These findings are broadly consistent with the aggregate visitation patterns previously reported by @Schneller2014-ry in the central Alps at Vorarlberg and Neumayer1999-il in the Austrian Alps ~55 km south of our study system, suggesting that our findings are at least regionally generalizeable. When considering these patterns from the perspective of pollination services to mountain flora, it is important to note that while bumble bees are the dominant Hymenopteran pollinators in our study system, flies (especially Syrphid hoverflies) are also abundant flower visitors in mountain ecosystems, and their visitation is mainly restricted to the open flower morphotypes (e.g. disc flowers) that receive less attention from bumble bees [@Neumayer1999-il]. In the context of our research question, we have emphasized the differences in tongue length between bumble bee species, but all bumble bees have relatively long tongues compared to the overall pollinator community, and the process of niche partitioning *among* bumble bee species is nested within the broader process of niche partitioning *between* bumble bees and other flower visitors.

The relationship between tongue length and floral preference was strong for funnel, bell, flag, and head flowers, which largely comports with the the established relationship between tongue length and corolla depth [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. Nevertheless, tongue length was a poor predictor of preference for disc, stalk-disc, and lip flowers, which agrees with studies demonstrating the insufficiency of simple morphological matching as an explanation of bumble bee foraging patterns [e.g. @Ranta1980-ok; @Goulson2008-oe]. Where tongue length fails as a predictor of floral preference, it can do so in two ways: the *convergence* of preference across species of *different* tongue length or the *divergence* of preference among species of *similar* tongue length. An example of the former is the near-universal affinity for lip flowers among the bumble bee species in our study. The only exception to this rule was the rare high-elevation species *B. mucidus*, which was effectively a flag flower specialist in our study. Lip flowers are known to be favored by bumble bees, particularly in the Alps [@Schneller2014-ry], and the apparent consistency of their attractiveness across tongue length differences might be due to their relatively permissive floral morphology (compared to flag or funnel flowers, for example) and/or their tendency to offer high per-inflorescence rewards compared to other floral morphotypes [@Neumayer1999-il]. Perhaps the most striking divergence of floral preference within tongue length class was the divisive effect of bell flowers on short-tongued bumble bees. *B. pyrenaeus*, *B. soroeensis*, *B. pratorum*, and *B. jonellus* visited bell flowers at high rates, while *B. terrestris/lucorum*, *B. psithyrus*, and *B. lapidarius* visited them relatively rarely. A similar split among the short-tongued species occurred with respect to head flowers, which were visited at high rates by *B. soroeensis* and *B. psithyrus* but at relatively low rates by other short-tongued species. There are no obvious morphological distinctions within short-tongued bumble bees that would account for these divergent affinities, and ecological drivers such as differences in spatial or temporal activity patterns are more likely explanations [@Sowig1989-ld; @Schneller2014-ry]. Nectar robbing is another mechanism by which the relationship between tongue length and corolla depth might be relaxed [@Brian1954-cy], but its influence in our data is unclear. Of the bumble bees in our study system, *B. terrestris/lucorum* and especially *B. wurflenii* are known to engage frequently in nectar robbing [@Brian1954-cy; @Neumayer1999-il]. In the case of *B. wurflenii*, however, its morphotype affinities did not differ markedly from those of other medium-tongued species, and it may be that the condition of medium-tongue length alone confers enough foraging flexibility in our study system to dampen the divergent effects of nectar robbing. Nectar robbing might, however, account for the deviation of *B. terrestris/lucorum* from most other short-tongued species with respect to flag flowers. With the exception of *B. lapidarius*, all short-tongued species besides *B. terrestris/lucorum* eschewed flag flowers, while *B. terrestris/lucorum* visited them at rates comparable to medium- and long-tongued species. Indeed, *B. terrestris/lucorum* was, on the whole, an eclectic forager, also demonstrating a relatively high affinity for the disc flowers that were largely ignored by other species. Together, these species-level idiosyncracies of preference within tongue length classes underscore the fact that factors other than the matching of tongue length and floral morphology --- such as affinities based on floral color, plant growth form, spatial distribution of floral patches, or habitat context [@Brian1957-dg; Harder1985-ls; @Sowig1989-ld] --- are at play in bumble bees' floral preferences, which may partly explain the coexistence of bumble bees of similar tongue lengths.

@@

While tongue length was a relatively strong (though imperfect) predictor of bumble bee species' morphotype preferences, it was a weak predictor of diet similarity (at the level of floral genus) across bumble bees species, echoing the findnings of @Durieux2000-uk. Diet ordination produced a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species, which is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. Nevertheless, the ordination explained less than 50% of total variance, and while short-tongued species separated clearly from long- and medium-tongued species, the latter overlapped extensively with each other. These results were corroborated by our HGAM analysis of niche overlap, where the only significant in overall niche overlap was between short- and long-tongued species. A caveat to these analyses, though, is that our bumble bee community included only 3 long-tongued species (*B. gerstaeckeri*, *B. hortorum*, and *B. mendax*), two of which (*B. gerstaeckeri* and *B. mendax*) were relatively rare. It is possible that clearer patterns of tongue-length-based diet differentiation would be seen in a more balanced community.

Community-mean diet overlap at the level of floral morphotype was consistently nearly twice as high as diet overlap at the level of floral genus. From a numerical perspective, this result is unsurprising simply because each morphotype comprises many genera. From an ecological perspective, though, this pattern suggests another answer to the question of how bumble bee species of similar tongue length can coexists: if tongue length primarily constrains bumble bee foraging at the coarse level of floral morphotype, floral diversity *within* morphotypes creates space for bumble bees with similar tongue lengths to avoid actual dietary overlap, thus mitigating interspecific competition.

@@

## Conclusion
Coexistence between bumble bee species with similar tongue lengths hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. In addition to addressing the question of bumble bee species coexistence, our results suggest how patterns of competition may respond to climate and land use change along mountain slopes, an issue of immediate concern for the conservation of bumble bees and their floral mutualists [@Sirois-Delisle2018-vi; @Soroye2020-zb].

This suggests that diversity within floral morphotypes may as important as morphotype diversity *per se* and overall floral abundance in facilitating the coexistence of bumble bee species, an insight which may have implications for the design of floral enhancements aimed at bumble bee conservation.

@@

<!-- Lip and head flowers have the highest nectar reward per inflorescence [@Neumayer1999-il] cited in . -->
<!-- Interestingly, @Farnberger1996-xo in a localized study near Wien found rather different results. Lip flowers were still found to be the favorite overall, and head flowers were popular, too, but flag flowers were of secondary importance, and bell flowers were the most visited morphotype for humi, pasc, and prat. Pasc almost totally ignored bell flowers in our study. -->
